Reasoning About a Machine with Local Capabilities

### 2018-05-03 toplas/paper.tex
Updated FTLR proof, malloc, and macro section with Lars' suggestions

In order to have a reasonably realistic system, we use a simple model of linking
where a program has access to a linking table that contains capabilities for
other programs. We also assume malloc to be part of the trusted computing base
satisfying a certain specification. Malloc and linking tables are described
further in the next section.
\itoplas{The specification of malloc is presented in \sectionname~\ref{sec:malloc-and-macros} as it uses the semantic model we build in Section~\ref{sec:logical-relation}.
For full details on the linking table, we refer to the technical
appendix~\citep{technical_appendix}.} \itoplas{(Reviewer A, esop suggest explaining the
  linking model)\\}
\itoplassug{Suggestion: We also have the flag table ussed
  for assertions, maybe we should add a description of it here.\\}
% \itoplassug{Reviewer C, popl, would have liked a discussion of the relationship
%   between this capability machines and old capability machines from the
%   litterature (Cambridge CAP \& IBM System 39, see tex comment).\\}
 % * I'm old, so I'd appreciate a brief discussion LCM's relationship
 %  with the capability hardware I (and perhaps other older readers, or
 %  those outside this precise subfield) are likely to have heard of ---
 %  the Cambridge CAP & IBM System 39.  I think the capabilities of the
 %  machine are not that far from CAP, but this machine has the local
 %  capabilities support. Is that right?

@@

\subsection{Safety of the Capability Machine}
With the logical relation defined, we can now state the fundamental theorem of
our logical relation: a strong theorem that formalizes the guarantees offered by
the capability machine. Essentially, it says a capability that only grants safe
authority is capability safe as a program counter.
\begin{theorem}[Fundamental Theorem]
  \label{thm:ftlr}
  If one of the following holds:
  \begin{align*}
      & \bullet
        \begin{aligned}[t]
        &\perm = \exec \text{ and }\npair{(\start,\addrend)} \in \readCond{}(\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwx \text{ and } \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and }\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\nwl,\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwlx \text{ and }\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and }\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\pwl,\gl)(W),
      \end{aligned}
  \end{align*}
  then $\npair{((\perm,\gl),\start,\addrend,\addr)} \in \stder(W)$
\end{theorem}
\begin{toplas}
  \begin{proof}[Proof sketch]
    Induction over $n$. By definition of $\stder(W)$, show
    $\npair[n']{(\reg\update{\pcreg}{((\perm,\gl),\start,\addrend,\addr)},\ms)}
    \in \observations(W)$ assuming $n' \leq n$, $\npair[n']{\reg} \in \stdrr(W)$, and
    $\heapSat[\ms]{n'}{W}$. By definition of $\observations$ let $\ms_f$,
    $\mem'$, and $i \leq n'$ be given and assume $\Phi \step[i]
    (\halted,\mem')$ and show there exists $W' \futurestr W$ that part of
    $\mem'$ satisfies. First observe that $i \neq 0$ as $\Phi$ is a non-halting
    configuration, so $\Phi$ takes at least one step, i.e.\ $\Phi \step
    \var{conf}' \step[i-1] (\halted,\mem')$.

    The rest of the proof roughly consists of two parts (1) argue that
    $\var{conf}'$ is consistent with the world in the sense that the
    register-file and memory still respect the world. (2) Argue that the rest of
    the execution respects the world. The second case, relies on where the
    $\pcreg$ in $\var{conf}'$ comes from. If the step to $\var{conf}'$ was a
    jump, then use the fact that the register-file contains safe values in the
    initial configuration. On the other hand, if the $\pcreg$ was just
    incremented, then we need to apply the induction hypothesis which is
    possible as all of the permission based conditions are downwards closed.
    For instance in the case where the $\pcreg$ is read-execute, i.e.\ $\perm =
    \exec$, we also assume $\npair{(\start,\addrend)} \in \readCond{}(\gl)(W)$
    which entails $\npair[n'-1]{(\start,\addrend)} \in \readCond{}(\gl)(W)$ and
    allows us to use the induction hypothesis.
    In order to argue (1), we just need to consider the configurations the
    operational semantics allows us to get to from the initial state. If we
    consider the memory in $\var{conf}'$, then it either (a) remains unchanged
    or (b) one address has been updated and in that case, the register-file contains an 
    appropriate capability for writing. The latter case, (b) corresponds to the first
    step executing a $\texttt{store}$ instruction, and the former (a) corresponds to
    any other instruction.
    In case (a), we can conclude that the memory still respects the world just
    by downwards closure of memory satisfaction. In the other case, (b), we use
    an auxiliary lemma called \emph{Conditions for \texttt{store} are
      sufficient} (see the appendix, Lemma~\ref{lem:conds-store-stuff})
    that states that given all the assumptions we have done so far, the updated
    memory still satisfies the world.
    All the possible changes to the register-file correspond to the effect of
    some instruction, so for the non-trivial instructions we use a sufficiency
    lemma.

@@

\begin{toplas}
  \section{Malloc and macros}
  \label{sec:malloc-and-macros}
With the semantic model defined, we would like to show its usefulness by proving
the correctness of a series of examples that rely
on well-bracketedness and local-state encapsulation and use the calling
convention to enforce these properties. However, the programs that
run on our capability are assembly programs and even program examples that would
be small in a high-level language, they become big and unintelligible at this
low level. In order to make our program examples intelligible, we introduce a
series of low-level abstractions in the form of macros.

To make the program examples realistic, they have access to a trusted malloc, so
they can allocate new memory. Our results are not specific to
any one implementation of malloc, so rather than providing the implementation,
we provide a specification that malloc must satisfy.

\subsection{Malloc specification}
  \itoplassug{Suggestion: add the malloc specification}
Before we give the malloc specification, we provide some intuition about its requirements.

@@

The requirements so far are syntactic and prevent direct tampering in the sense
that the capability intended for invoking malloc only can be used for
invocation. This does, however, not prevent the implementation of malloc from
colluding with other code. Specifically, malloc could depend on the state of a
piece of memory not just under the control of malloc. We also need to make sure
that malloc actually behave as intended. That is when malloc is invoked with a
length argument, then it actually returns a capability for a fresh piece of
memory of that size. It should be fresh in the sense that malloc hasn't given
out a capability for any part of that memory before and won't do it in the
future. Finally, we also need to make sure that malloc behaves correctly when
given a nonsensical length argument such as a negative integer or a capability.
These requirements relate to the behaviour of malloc, so we cannot hope to
express them syntactically. Instead we will use the semantic model defined in
Section~\ref{sec:logical-relation} to express this part of the specification.

In order to express the semantic requirements on malloc, the specification will
require the existence of some initial region for malloc. Malloc does not
disappear from the system, so the malloc region must be permanent. This also
means that the malloc specification should guarantee that use of malloc is safe
at any point of execution no matter who has invoked it. To this end, the
requirement on the initial malloc region should also hold true in any private
future world.
% priv future means that malloc should not depend on local capabilities?

The implementation of malloc should not depend on any memory outside the
implementation of malloc. This means that malloc should not have any
capabilities in its internals that govern other programs. As explained in
\sectionname~\ref{subsec:worlds}, a region's state interpretation function
determines what memory segments it permits with respect to some world. The state
interpretation function of malloc should permit the same regions with respect to
a world with the malloc region as it would with respect to a world that only
contains the malloc region. In other words, the remaining regions in the world
do not influence which regions malloc accepts. Malloc should retain this
property throughout execution, so it must hold true for any private future
region of the initial malloc region.

The malloc specification will also dictate what malloc should do when invoked.
Malloc can be invoked with either a valid or an invalid length argument. In the
case where malloc is invoked with an invalid argument, that is a negative
integer or a capability, we require malloc to simply fail. This part of the
specification does not rely on the malloc region as failing is considered
safe behaviour. When malloc is invoked with a valid length, then malloc should
return a fresh memory segment of the correct length.
\todo[inline]{Everything below this is work in progress..}
Malloc may have been invoked a number of times before any given invocation, so
the internal data structures of malloc may be in a different state than it
started in.
We know that the memory of malloc is in a state consistent with the
protocol dictated by the initial malloc region in the sense that we can assume
that malloc is invoked in a configuration where part of it satisfies a private
future region of the initial malloc region. Knowing the state of the malloc
region allows us to express what we mean by a \emph{fresh} region. This means
that malloc should not keep any capabilities for the freshly allocated region.
To ensure this, we require the footprint of malloc after the call to be disjoint
from the freshly allocated memory. Further, the footprint should satisfy a
public future region of the malloc region the invocation happened in. This means
that it cannot depend on the freshly allocated region. It further allows
subsequent calls to malloc as it provides one of the requirements in the malloc
specification.

Malloc is part of the trusted computing base. If we cannot trust malloc, then we
cannot use the memory it allocates as we have no idea whether it is aliased by
some untrusted program. Because of this assumption, we also require malloc to
promise that the non-argument registers of the register-file remain unchanged
after the call. This allows the caller to keep private capabilities in the
register file. This is strictly speaking not necessary as we can use our calling
convention to protect our local state.

\begin{definition}[Malloc Specification]
  \label{spec:malloc}
  $c_\malloc$ satisfies the specification for malloc iff
  \[  
    \begin{array}{l}
      c_\malloc = ((\entry,\glob),\_,\_,\_) \land\\
      \exists \iota_{\malloc,0} \ldotp \\
      \quad \iota_{\malloc,0}.v = \perma \; \land \\
      \quad \left(
        \begin{array}{l}
          \forall \iota' \futurestr \iota_{\malloc,0}, W,i \ldotp\\
          \quad W(i)=\iota' \Rightarrow \iota'.H (\iota'.s) (\xi^{-1}(W)) = \iota'.H (\iota'.s) (\xi^{-1}([i \mapsto W(i)]))
        \end{array}
        \right) \wedge \\
      \quad \left(
      \begin{array}{l}
        \forall \Phi \in \ExecConfs \ldotp \forall \ms_{\var{footprint}}, \hsframe \in \HeapSegments \ldotp \\
        \quad \forall i, n, \size \in \nats \ldotp \forall
        w_{\var{ret}} \in \Words \ldotp \\
        \quad \quad \forall \iota_\malloc \futurestr \iota_{\malloc,0} \land \\
        \quad \quad \memheap = \ms_{\var{footprint}} \uplus \hsframe \land \heapSat[\ms_{\var{footprint}}]{n}{[i \mapsto \iota_\malloc]} \land \\
        \qquad \memreg(r_1) = \size \land \size \geq 0 \land  \memreg(r_0) = w_{\var{ret}} \land \\
        \qquad \memreg(\pcreg) = \updatePcPerm{c_\malloc} \\ %and this needs to change with ABI.
        \qquad \Rightarrow \\
        \qquad \quad \exists \Phi' \in \ExecConfs \ldotp \exists \ms_{\var{footprint}}', \ms_{\var{alloc}} \in \HeapSegments\ldotp\\
        \qquad \qquad \exists j \in \nats \ldotp j > 0 \land \exists b',e'\in \Addrs \ldotp \exists \iota_\malloc' \in \Regions \ldotp \\
        \qquad \qquad \quad \Phi \step[j] \Phi' \land \\
        \qquad \qquad \quad \memheap[\Phi']=\ms_{\var{footprint}}' \uplus \hs_{\var{alloc}} \uplus \hsframe \land\\
        \qquad \qquad \quad \iota_{\malloc}' \futurewk \iota_\malloc \land \\
        \qquad \qquad \quad \heapSat[\ms_{\var{footprint}}']{n-j}{[i \mapsto \iota_\malloc']} \land \\
        \qquad \qquad \quad \dom(\hs_{\var{alloc}}) = [b',e'] \land \forall a \in [b',e']\ldotp \hs_{\var{alloc}}(a) = 0  \land \\
        \qquad \qquad \quad \memreg[\Phi'] = \memreg[\Phi]\update{\pcreg}{\updatePcPerm{w_{\var{ret}}}}\update{r_1}{((\rwx,\glob),b',e',b')} \land \\
        \qquad \qquad \quad \size - 1 = e'-b'
      \end{array}
      \right) \land\\
      \quad \left(
      \begin{array}{l}
        \forall \Phi \in \ExecConfs \ldotp\\
        \quad(\memreg(r_1) \notin \ints \vee \memreg(r_1) < 0) \land
        \memreg(\pcreg) = \updatePcPerm{c_\malloc} \\
        \qquad \Rightarrow \exists j \in \nats \ldotp \Phi \step[j] \failed
      \end{array}
        \right)
    \end{array}
  \]
\end{definition}
As described previously, the specification of malloc ensure that malloc has no
capabilities pointing \emph{out} of malloc. It does, however, not say anything about
capabilities that point \emph{in} to malloc. We obviously can obviously not
allow this if we want to trust malloc. We have chosen to keep the malloc
specification simple and let this assumption be in the lemmas that use malloc.
It is sufficient for these lemmas to require that there are no outside
capabilities for malloc in the initial configuration as capabilities cannot
appear out of thin air and the malloc specification makes sure that capabilities
are not leaked.

Malloc should not just be available to trusted programs, but also to possibly
malicious programs. The specification should therefore entail that the malloc
capability is safe which would make it safe to let even malicious programs have
access to it.
\begin{lemma}[Malloc is safe to pass to adversary]
  \label{lem:malloc-in-vr}
  For $c_\malloc$ that satisfies the specification for malloc with region $\iota_{\malloc,0}$,  if $W(r) \futurestr \iota_{\malloc,0}$, then
  $\npair{c_\malloc} \in \stdvr(W)$ for all $n$.
\end{lemma}

\subsection{Reusable macro instructions}
\label{subsec:macros}
The focus of this work is not to give a definitive specification and
implementation of malloc which is why we do give a malloc implementation. We
are, however, confident that it is possible to make an implementation of malloc
that satisfies the malloc specification in Definition~\ref{spec:malloc}. There
are in particular two simplifications in our system that makes things easy:
First, we do not consider deallocation of memory which means that the data
structure malloc uses to keep track of free memory does not have to handle
reclaimed memory. Second, the malloc specification does nor permit malloc to run
out of memory and thus prevent allocation. This is possible on our simple
capability machine because it has an infinite address space. The capability with
an infinite range of authority would of course need to be part of malloc, but it
could also double as the data structure that keeps track of free memory.
\itoplassug{If we have references for malloc being difficult, we could add that
  to the text.}
%\itoplassug{Add the assumption that the malloc capability in examples
%  satisfies specification.}

%\itoplassug{Add examples of macros to this section to make it clear what we mean by macros.\\}
%\itoplassug{Reviewer C, ESOP, asks for high-level specifications for the macros in the style of malloc. I am not sure we want to actually include this in the submission, so it is maybe not here we should put our attention. (see review in tex comment)\\}
% Finally, as already mentioned above, the approach relies on `call` and `scall` macro-instructions that are only outlined in the submission, and described by very complicated pseudocode and English prose in the companion technical appendix.  I still miss a fully formal description of 1- what instructions the macros expand to, exactly, and 2- what contract they satisfy (in the style of the contract for malloc in section 2 of the technical appendix).
We define a number of reusable macros capturing the calling convention as well
as other conveniences. The macros that utilise the stack assume that it is
available in register $r_\stk$.
\itoplassug{Lau, suggestion: should we mention temporary registers? They are
  used in the macros, but it seems like a minor technical detail.}

The macro \texttt{\footnotesize{scall
    $r$($\overline{r_{\var{args}}}$,$\overline{r_{\var{priv}}}$)}} captures the
parts of the calling convention related to actually transferring control to some
adversarial code. Specifically, it pushes the contents of the private registers,
$\overline{r_{\var{priv}}}$, to the stack. Then it pushes the restoration code
to the stack. The restoration code will be executed as the first thing upon
return, where it restores the stack pointer and the old program pointer. After
the restoration code is pushed to the stack, a copy of the pc is adjusted to
point to the first instruction after the control has been transferred and then
pushed to the stack. From the stack pointer, a protected return pointer is
created by adjusting it to point to the return code and encapsulating it by
restricting its permission to enter. Next, the stack pointer is restricted to
the unused part, and the unused part of the stack is cleared, so it does not
govern any part of the private stack frame which means that it can safely be
given away. Finally, the non-argument registers are cleared and we jump to $r$.
Upon return after the restoration code has been executed, the restoration code
is popped from the stack and the private words we pushed to the stack before the
call are popped to the private registers. The implementation of \texttt{scall}
is presented in \figurename~\ref{fig:scall-impl}, and the restoration code is
presented in \figurename~\ref{fig:scall-rest-code}. The implementation of
\texttt{scall} uses some of the macros we present later in this section.

The macros \texttt{\footnotesize{prepstack $r$}} and \texttt{\footnotesize{reqglob
    $r$}} are the last macros related to the calling convention. The former,
\texttt{prepstack} should be used when one receives a stack from an unknown
sure. The \texttt{prepstack} macro first ensures the stack has permission
read/write-local/execute, and then it adjusts the stack capability, so it
follows the convention for the stack\footnote{The stack capability should always
point to the top most word on the stack. A stack received from an unknown source
can be treated as empty, so the stack capability should point just outside its
range of authority.}. The other macro, \texttt{reqglob}, ensures that the
capability in register $r$ is global. This macro should be used on callbacks
received from unknown sources to ensure that they cannot be derived from the
stack pointer.

The remaining macros are not directly related to the calling convention, but
they help making the examples in \sectionname~\ref{sec:examples} intelligible.
The macros \texttt{\footnotesize{push $r$}} and \texttt{\footnotesize{pop $r$}}
add and remove elements from the stack. The macro \texttt{\footnotesize{fetch
    $r$ $\var{name}$}} fetches the capability related to $\var{name}$ from the
linking table and stores it in register $r$. The macro
\texttt{\footnotesize{malloc $r$ $n$}} invokes malloc with size argument $n$.
The \texttt{malloc} macro assumes that a capability for malloc resides in the
linking table and is basically a fetch of the malloc capability followed by a
setup of a return pointer. Finally, the macro \texttt{\footnotesize{crtcls
    $\overline{(x_i,r_i)}$ $r$}} allocates a closure where $r$ points to the
closure's code and a new environment is allocated (using malloc) where the
contents of $\overline{r_i}$ is stored. In the code referred to by $r$, an
implicit fetch happens when an instruction refers to $x_i$.

The Appendix contains the implementation of all the macros used in
\texttt{scall}. The technical appendix~\citep{technical_appendix} contains detailed descriptions
of all the macros as well as all implementations.
%\itoplassug{Suggestion: Add the call implementation (and possibly other macros?) in the above?}

@@

>>>>

In order to have a reasonably realistic system, we use a simple model of linking
where a program has access to a linking table that contains capabilities for
other programs. We also assume malloc to be part of the trusted computing base
satisfying a certain specification. Malloc and linking tables are described
further in the next section.
\itoplas{The specification of malloc is presented in \sectionname~\ref{sec:malloc} as it uses the semantic model we build in Section~\ref{sec:logical-relation}.
For full details on the linking table, we refer to the technical
appendix~\citep{technical_appendix}.} \itoplas{(Reviewer A, esop suggest explaining the
  linking model)\\}
\itoplassug{Suggestion: We also have the flag table ussed
  for assertions, maybe we should add a description of it here.\\}
% \itoplassug{Reviewer C, popl, would have liked a discussion of the relationship
%   between this capability machines and old capability machines from the
%   litterature (Cambridge CAP \& IBM System 39, see tex comment).\\}
 % * I'm old, so I'd appreciate a brief discussion LCM's relationship
 %  with the capability hardware I (and perhaps other older readers, or
 %  those outside this precise subfield) are likely to have heard of ---
 %  the Cambridge CAP & IBM System 39.  I think the capabilities of the
 %  machine are not that far from CAP, but this machine has the local
 %  capabilities support. Is that right?

@@

\subsection{Safety of the Capability Machine}
With the logical relation defined, we can now state the fundamental theorem of
our logical relation: a strong theorem that formalizes the guarantees offered by
the capability machine. Essentially, it says a capability that only grants safe
authority is capability safe as a program counter.
\begin{theorem}[Fundamental Theorem]
  \label{thm:ftlr}
  If one of the following holds:
  \begin{align*}
      & \bullet
        \begin{aligned}[t]
        &\perm = \exec \text{ and }\npair{(\start,\addrend)} \in \readCond{}(\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwx \text{ and } \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and }\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\nwl,\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwlx \text{ and }\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and }\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\pwl,\gl)(W),
      \end{aligned}
  \end{align*}
  then $\npair{((\perm,\gl),\start,\addrend,\addr)} \in \stder(W)$
\end{theorem}
\begin{toplas}
  \begin{proof}[Proof sketch]
    Induction over $n$. By definition of $\stder(W)$, show
    $\npair[n']{(\reg\update{\pcreg}{((\perm,\gl),\start,\addrend,\addr)},\ms)}
    \in \observations(W)$ assuming $n' \leq n$, $\npair[n']{\reg} \in \stdrr(W)$, and
    $\heapSat[\ms]{n'}{W}$. By definition of $\observations$ let $\ms_f$,
    $\mem'$, and $i \leq n'$ be given and for
    $\Phi = (\reg\update{\pcreg}{((\perm,\gl),\start,\addrend,\addr)},\ms \uplus
    \ms_f)$ assume $\Phi \step[i]
    (\halted,\mem')$ and show there exists $W' \futurestr W$ that part of
    $\mem'$ satisfies. First observe that $i \neq 0$ as $\Phi$ is a non-halting
    configuration, so $\Phi$ takes at least one step, i.e.\ $\Phi \step
    \Phi' \step[i-1] (\halted,\mem')$.

    The rest of the proof roughly consists of two parts: (1) Argue that $\Phi'$
    is consistent with the world in the sense that the register-file and memory
    still respect the world. (2) Argue that the rest of the execution respects
    the world. The second case relies on where the $\pcreg$ in $\Phi'$ comes
    from. If the step to $\Phi'$ was a jump, then use the fact that the
    register-file contains safe values in $\Phi$. On the other hand, if the
    $\pcreg$ was just incremented, then we need to apply the induction
    hypothesis, which is possible as all of the permission based conditions are
    downwards closed. For instance, in the case where the $\pcreg$ is
    read-execute, i.e.\ $\perm = \exec$, we also assume
    $\npair{(\start,\addrend)} \in \readCond{}(\gl)(W)$ which entails
    $\npair[n'-1]{(\start,\addrend)} \in \readCond{}(\gl)(W)$ and hence allows us to
    use the induction hypothesis. In order to argue (1), we just need to
    consider the configurations the operational semantics allows us to get to
    from the initial state. If we consider the memory in $\Phi'$, then it either
    (a) remains unchanged or (b) one address has been updated and in that case,
    the register-file contains an appropriate capability for writing. The latter
    case (b) corresponds to the first step executing a $\texttt{store}$
    instruction, and the former (a) corresponds to any other instruction. In
    case (a), we can conclude that the memory still respects the world just by
    downwards closure of memory satisfaction. In the other case, (b), we use an
    auxiliary lemma called \emph{Conditions for \texttt{store} are sufficient}
    (see the appendix, Lemma~\ref{lem:conds-store-stuff}) that states that given
    all the assumptions we have so far, the updated memory still satisfies
    the world. All the possible changes to the register-file correspond to the
    effect of some instruction, so for each non-trivial instruction we use a
    lemma appropriate for that instruction.

@@

\begin{toplas}
  \section{Malloc}
  \label{sec:malloc}
In the examples we present in \sectionname~\ref{sec:examples}, we will assume
the existence of a trusted malloc routine so that it is possible to allocate new
memory. Our semantic model is not specific to a particular implementation of
malloc, so rather than providing the implementation, we provide a specification
that malloc must satisfy. This section can be skipped on first read through.

@@

The requirements so far are syntactic and prevent direct tampering in the sense
that the capability intended for invoking malloc only can be used for
invocation. This does, however, not prevent the implementation of malloc from
colluding with other code. Specifically, malloc could depend on the state of a
piece of memory not just under the control of malloc. We also need to make sure
that malloc actually behaves as intended, that is, when malloc is invoked with a
length argument, then it actually returns a capability for a fresh piece of
memory of that size. It should be fresh in the sense that malloc has not given
out a capability for any part of that memory before and will not do it in the
future. Finally, we also need to make sure that malloc behaves correctly when
given a nonsensical length argument such as a negative integer or a capability.
These requirements relate to the behaviour of malloc, so we cannot hope to
express them syntactically. Instead we will use the semantic model defined in
Section~\ref{sec:logical-relation} to express this part of the specification.

For the semantic requirements on malloc, the specification will
require the existence of some initial region for malloc. Malloc does not
disappear from the system, so the malloc region must be permanent. This also
means that the malloc specification should guarantee that use of malloc is safe
at any point of execution no matter who has invoked it. To this end, the
requirement on the initial malloc region should also hold true in any private
future world.
% priv future means that malloc should not depend on local capabilities?

The implementation of malloc should not depend on any memory outside the
implementation of malloc. This means that malloc should not have any
capabilities in its internals that govern other programs. As explained in
\sectionname~\ref{subsec:worlds}, a region's state interpretation function
determines what memory segments it permits with respect to some world. The state
interpretation function of malloc should permit the same regions with respect to
a world with the malloc region as it would with respect to a world that only
contains the malloc region. In other words, the remaining regions in the world
should not influence which regions malloc accepts. Malloc should retain this
property throughout execution, so it must hold true for any private future
region of the initial malloc region.

The malloc specification will also dictate what malloc should do when invoked.
Malloc can be invoked with either a valid or an invalid length argument. In the
case where malloc is invoked with an invalid argument, that is, a negative
integer or a capability, we simply require malloc to fail. This part of the
specification does not rely on the malloc region, as failing is considered
safe behaviour. When malloc is invoked with a valid length, then malloc should
return a fresh memory segment of the correct length.
Malloc may have been invoked a number of times before any given invocation, so
the internal data structures of malloc may be in a different state than they
were initially.
We know that the memory of malloc is in a state consistent with the
protocol dictated by the initial malloc region in the sense that we can assume
that malloc is invoked in a configuration where part of it satisfies a private
future region of the initial malloc region. Knowing the state of the malloc
region allows us to express what we mean by a \emph{fresh} region. This means
that malloc should not keep any capabilities for the freshly allocated region.
To ensure this, we require the footprint of malloc after the call to be disjoint
from the freshly allocated memory. Further, the footprint should satisfy a
public future region of the malloc region the invocation happened in. This means
that it cannot depend on the freshly allocated region. It further allows
subsequent calls to malloc as it provides one of the requirements in the malloc
specification.

Malloc is part of the trusted computing base. If we cannot trust malloc, then we
cannot use the memory it allocates as we have no idea whether it is aliased by
some untrusted program. Because of this assumption, we also require malloc to
promise that the non-argument registers of the register-file remain unchanged
after the call. This allows the caller to keep private capabilities in the
register file (strictly speaking this is not necessary as we can use our calling
convention to protect our local state).

\begin{definition}[Malloc Specification]
  \label{spec:malloc}
  $c_\malloc$ satisfies the specification for malloc iff
  \[  
    \begin{array}{l}
      c_\malloc = ((\entry,\glob),\_,\_,\_) \land\\
      \exists \iota_{\malloc,0} \ldotp \\
      \quad \iota_{\malloc,0}.v = \perma \; \land \\
      \quad \left(
        \begin{array}{l}
          \forall \iota' \futurestr \iota_{\malloc,0}, W,i \ldotp\\
          \quad W(i)=\iota' \Rightarrow \iota'.H (\iota'.s) (\xi^{-1}(W)) = \iota'.H (\iota'.s) (\xi^{-1}([i \mapsto W(i)]))
        \end{array}
        \right) \wedge \\
      \quad \left(
      \begin{array}{l}
        \forall \Phi \in \ExecConfs \ldotp \forall \ms_{\var{footprint}}, \hsframe \in \HeapSegments \ldotp \\
        \quad \forall i, n, \size \in \nats \ldotp \forall
        w_{\var{ret}} \in \Words \ldotp \\
        \quad \quad \forall \iota_\malloc \futurestr \iota_{\malloc,0} \land \\
        \quad \quad \memheap = \ms_{\var{footprint}} \uplus \hsframe \land \heapSat[\ms_{\var{footprint}}]{n}{[i \mapsto \iota_\malloc]} \land \\
        \qquad \memreg(r_1) = \size \land \size \geq 0 \land  \memreg(r_0) = w_{\var{ret}} \land \\
        \qquad \memreg(\pcreg) = \updatePcPerm{c_\malloc} \\ %and this needs to change with ABI.
        \qquad \Rightarrow \\
        \qquad \quad \exists \Phi' \in \ExecConfs \ldotp \exists \ms_{\var{footprint}}', \ms_{\var{alloc}} \in \HeapSegments\ldotp\\
        \qquad \qquad \exists j \in \nats \ldotp j > 0 \land \exists b',e'\in \Addrs \ldotp \exists \iota_\malloc' \in \Regions \ldotp \\
        \qquad \qquad \quad \Phi \step[j] \Phi' \land \\
        \qquad \qquad \quad \memheap[\Phi']=\ms_{\var{footprint}}' \uplus \hs_{\var{alloc}} \uplus \hsframe \land\\
        \qquad \qquad \quad \iota_{\malloc}' \futurewk \iota_\malloc \land \\
        \qquad \qquad \quad \heapSat[\ms_{\var{footprint}}']{n-j}{[i \mapsto \iota_\malloc']} \land \\
        \qquad \qquad \quad \dom(\hs_{\var{alloc}}) = [b',e'] \land \forall a \in [b',e']\ldotp \hs_{\var{alloc}}(a) = 0  \land \\
        \qquad \qquad \quad \memreg[\Phi'] = \memreg[\Phi]\update{\pcreg}{\updatePcPerm{w_{\var{ret}}}}\update{r_1}{((\rwx,\glob),b',e',b')} \land \\
        \qquad \qquad \quad \size - 1 = e'-b'
      \end{array}
      \right) \land\\
      \quad \left(
      \begin{array}{l}
        \forall \Phi \in \ExecConfs \ldotp\\
        \quad(\memreg(r_1) \notin \ints \vee \memreg(r_1) < 0) \land
        \memreg(\pcreg) = \updatePcPerm{c_\malloc} \\
        \qquad \Rightarrow \exists j \in \nats \ldotp \Phi \step[j] \failed
      \end{array}
        \right)
    \end{array}
  \]
\end{definition}
As described previously, the specification of malloc ensures that malloc has no
capabilities pointing \emph{out} of malloc. It does, however, not say anything about
capabilities that point \emph{in} to malloc. We obviously can not
allow this if we want to trust malloc. We have chosen to keep the malloc
specification simple and let this assumption be in the lemmas that use malloc.
It is sufficient for these lemmas to require that there are no outside
capabilities for malloc in the initial configuration as capabilities cannot
appear out of thin air and the malloc specification makes sure that capabilities
are not leaked.

Malloc should not just be available to trusted programs, but also to possibly
malicious programs. The specification should therefore entail that the malloc
capability is safe which would make it safe to let even malicious programs have
access to it.
\begin{lemma}[Malloc is safe to pass to adversary]
  \label{lem:malloc-in-vr}
  For all $c_\malloc$ that satisfies the specification for malloc with region $\iota_{\malloc,0}$,  if $W(r) \futurestr \iota_{\malloc,0}$, then
  $\npair{c_\malloc} \in \stdvr(W)$ for all $n$.
\end{lemma}

The focus of this work is not to give a definitive specification and
implementation of malloc which is why we do give a malloc implementation. We
are, however, confident that it is possible to make an implementation of malloc
that satisfies the malloc specification in Definition~\ref{spec:malloc}. There
are in particular two simplifications in our system that makes things easy:
First, we do not consider deallocation of memory which means that the data
structure malloc uses to keep track of free memory does not have to handle
reclaimed memory. Second, the malloc specification does nor permit malloc to run
out of memory and thus prevent allocation. This is possible on our simple
capability machine because it has an infinite address space. The capability with
an infinite range of authority would of course need to be part of malloc, but it
could also double as the data structure that keeps track of free memory.
\itoplassug{If we have references for malloc being difficult, we could add that
  to the text.}
%\itoplassug{Add the assumption that the malloc capability in examples
%  satisfies specification.}

\section{Reusable macro instructions}
\label{sec:macros}
%\itoplassug{Add examples of macros to this section to make it clear what we mean by macros.\\}
%\itoplassug{Reviewer C, ESOP, asks for high-level specifications for the macros in the style of malloc. I am not sure we want to actually include this in the submission, so it is maybe not here we should put our attention. (see review in tex comment)\\}
% Finally, as already mentioned above, the approach relies on `call` and `scall` macro-instructions that are only outlined in the submission, and described by very complicated pseudocode and English prose in the companion technical appendix.  I still miss a fully formal description of 1- what instructions the macros expand to, exactly, and 2- what contract they satisfy (in the style of the contract for malloc in section 2 of the technical appendix).
With the semantic model defined, we would like to show its usefulness by proving
the correctness of a series of examples that rely
on well-bracketedness and local-state encapsulation and use the calling
convention to enforce these properties. However, the programs that
run on our capability are assembly programs and even program examples that would
be small in a high-level language become big and unintelligible at this
low level. Thus to make our program examples intelligible, we introduce a
series of low-level abstractions in the form of macros.
We define a number of reusable macros capturing the calling convention as well
as other conveniences. The macros that utilise the stack assume that it is
available in register $r_\stk$.
\itoplassug{Lau, suggestion: should we mention temporary registers? They are
  used in the macros, but it seems like a minor technical detail.}

The macro \texttt{\footnotesize{scall
    $r$($\overline{r_{\var{args}}}$,$\overline{r_{\var{priv}}}$)}} captures
those parts of the calling convention related to actually transferring control
to some adversarial code. Specifically, it pushes the contents of the private
registers, $\overline{r_{\var{priv}}}$, to the stack. Then it pushes the
``restoration'' code to the stack. The restoration code will be executed as the
first thing upon return, and restores the stack pointer and the old program
pointer. After the restoration code is pushed to the stack, a copy of the pc is
pushed to the stack after it has been adjusted to point to the first instruction
after the jump. From the stack pointer, a protected return pointer is created by
adjusting it to point to the return code and encapsulating it by restricting its
permission to Enter. Next, the stack pointer is restricted to the unused part,
and the unused part of the stack is cleared, so it does not govern any part of
the private stack frame which means that it can safely be given away. Finally,
the non-argument registers are cleared and we jump to $r$. Upon return after the
restoration code has been executed, the restoration code is popped from the
stack and the private words we pushed to the stack before the call are popped to
the private registers. The implementation of \texttt{scall} is presented in
\figurename~\ref{fig:scall-impl}, and the restoration code is presented in
\figurename~\ref{fig:scall-rest-code}. The implementation of \texttt{scall} uses
some of the macros we now present.

@@

The macros \texttt{\footnotesize{prepstack $r$}} and \texttt{\footnotesize{reqglob
    $r$}} are the last macros related to the calling convention. The former,
\texttt{prepstack} should be used when one receives a stack from an unknown
source. The \texttt{prepstack} macro first ensures the stack has permission
read/write-local/execute, and then it adjusts the stack capability, so it
follows the convention for the stack\footnote{The stack capability should always
point to the top most word on the stack. A stack received from an unknown source
can be treated as empty, so the stack capability should point just outside its
range of authority.}. The other macro, \texttt{reqglob}, ensures that the
capability in register $r$ is global. This macro should be used on callbacks
received from unknown sources to ensure that they cannot be derived from the
stack pointer.
